\chapter{Introduction}

Scientific computing has evolved from being essentially the only kind of
computing that existed, to being a small part of how and why computers
are used.
Over this period of time, programming tools have advanced in terms of
the abstractions and generalizations they are capable of.

Science as a whole evolves through the incorporation of specialized bits
of knowledge into more powerful general theories.  
There is a parallel in the world of programming languages:
special-purpose languages have been subsumed by more powerful and general
languages.
How has this trend affected scientific computing?
The surprising answer is: not as much as we would like.
We see scientists and technical users generally either turning away
from the best new programming languages, or else pushing them
to their limits in unusual ways.

%% Much of the history of programming languages has been about increasing abstraction
%% and generalization. Today we take it for granted that a single suitably powerful
%% language could be used for nearly any programming task, were it not for pragmatic
%% factors beyond our control. Over time,
%% The field of technical computing --- programming for applied math and the
%% sciences --- is something of a holdout from this trend, with its own
%% purpose-built languages still dominating.
%% There are probably many reasons for this, including historical, social and
%% technical.

\begin{singlespace}
\begin{figure}
  \begin{center}
    \begin{tabular}{|llll|l|}\hline
      \multicolumn{4}{|c|}{Productivity} & Performance \\
      \hline
      Matlab  &  Maple &  Mathematica & SciPy & Fortress\\
      SciLab  &  IDL   &  R  & Octave         & Chapel \\
      S-PLUS  & SAS & J & APL                 & X10 \\
      Maxima & Mathcad & Axiom & Sage         & SAC \\
      Lush & Ch & LabView & O-Matrix          & ZPL \\
      PV-WAVE & Igor Pro & OriginLab & FreeMat &\\
      Yorick & GAUSS & MuPad & Genius &\\
      SciRuby & Ox & Stata & JLab &\\
      Magma & Euler & Rlab & Speakeasy &\\
      GDL & Nickle & gretl & ana &\\
      Torch7 & A+ & PDL & Nial & \\
      \hline
    \end{tabular}
  \end{center}
  \caption[49 technical computing languages]{
\small{
    49 technical computing languages classified as primarily designed for productivity or performance
}
  }
  \label{gangof40}
\end{figure}
\end{singlespace}

In fact, we have discovered at least \emph{49} programming languages
designed for technical computing since Fortran (figure~\ref{gangof40}).
This high number begs for an explanation.
We propose that technical users crave the flexibiltiy to pick notation
for their problems, and
language design is where you go when you need this level of flexibility.
Language design is associated with the highest level of abstraction.
The desire for scientists to rework their code is not so surprising when one reflects
on the tradition of inventing new notation in mathematics and science.
%This certainly
%helps distinguish scientific programming from other application areas.

The need for language-level flexibility is corroborated by
the ways that the technical computing domain uses general purpose
languages.
Effective scientific libraries extensively employ
polymorphism, custom operators, and compile time abstraction.
Code generation approaches (writing a program that writes the needed
program) are unusually common.
Most of these techniques are presently fairly difficult to use, and so
programmers working at this level give up the notational conveniences
of the purpose-built languages above.

%Yet it also demands the greatest ease-of-use.
%A good rule of thumb is that if you need to write a program that prints
%out code in language X, then language X is not powerful enough.

%pathologies arising from the language level are subtle
%not as immediate as ``this new matrix multiply is faster''

This thesis addresses the essential challenges created by
this state of affairs.
We argue that the situation   
can be improved considerably by
abstracting, with a new language.

%the essential challenges from the state
%f affairs we have described.
%Of course, a language is not the only thing the world of technical
%computing needs.
Compiler techniques, library design, high-performance
computational kernels, new algorithms, and approaches to parallelism are
all important to technical computing.
%ultimately be more important than language design.
However these sorts of technologies can usually be applied to multiple
languages, as has happened in the C and Fortran language families.
However in conjunction with a new language, these technologies can flourish.
Indeed, we believe that language design
has reached a level of importance that even (perhaps temporarily) supersedes
the traditional focus of cache utilization and the use of matrix operations.

%So in our view languages targeting this space need to dig deeper,
%at least temporarily forgetting about matrices and cache utilization,
%and look for bigger patterns.

%\todo[inline]{{\it Please consider if this is even more true: }\newline
%Technical computing  can benefit from many technologies.
%Some of these include compiler techniques, library design, high-performance
%computational kernels, new algorithms, and new approaches to parallelism.
%These sorts of technologies can usually be applied to multiple
%languages, as has happened in the C and Fortran language families.
%They do not by themselves insist that a new language be developed.
%However in conjunction with a new language, these technologies can flourish.
%}

In brief, our idea is to \emph{integrate code selection and specialization}.
The aforementioned flexibility requirement can be explained as
a sufficiently powerful means for \emph{selecting} which piece of code
to run.
This notion subsumes method dispatch, function overloading,
and potentially even branches.
When such mechanisms are used in technical computing, there is nearly
always a corresponding need to \emph{specialize} code for specific cases
to obtain performance.
Polymorphic languages sometimes support forms of specialization,
but often only through a designated mechanism (e.g. templates), or
deployed rather conservatively out of resource concerns.
We intend to show that an equilibrium point in the design space can be found
by combining selection and specialization into a single dynamic type-based
dispatch mechanism.
The specific design is identified via subtyping theories and closure under
data-flow operations.

%what does this system achieve?
%-

% why dynamic dispatch?
%% somewhat counter-intuitively, dynamic dispatch can be good for performance
%% since it permits invoking the most specialized possible method.
%% static overloading can lead to calling a sub-optimal case when multiple
%% overloads exist for the sake of performance.


%Extensive code specialization is a key feature of
%technical computing. ``what to specialize on'' has been an open problem.
%our types are a possible solution to this for two reasons:

%\begin{enumerate}
%\item you can tune the amount of information they contain
%\item everybody agrees to use them, which helps ensure that specializing on
%  the types will actually do something useful.
%\end{enumerate}

%the amazing thing about programming languages it that a better
%explanation can directly lead to better performance!)


%By \emph{selection} we mean any mechanism used to pick one of several
%pieces of code to run. This includes object-oriented mechanisms, as well
%as function overloading, and even branches.

%By \emph{specialization} we mean transformation of a given piece of code,
%particularly as done by a compiler to generate efficient code for some
%special case of a flexible piece of high-level code.

%% Specialization entails program analysis. By necessity, a language's
%% code selection mechanisms must inform this process, to ensure the
%% correctness of the analysis. A central argument of this thesis is that
%% specialization should feed back into selection, allowing the
%% \emph{approximate values} processed by a compiler to be used in the
%% source language for dispatch. We argue that this is the key feature
%% missing from the present generation of dynamically typed languages.
%This conclusion would not be surprising to advocates of static
%type systems, but the approach we propose is actually a run-time
%mechanism that does not restrict the class of valid programs.

%Why integrate code specialization and code selection?
%\begin{itemize}
%\item specialization requires selection anyway
%\item simpler system
%\item can sometimes collapse 2 layers of dispatch into 1
%\item can replace library code with a code generator without either changing
%  client code *or* any extra overhead
%\end{itemize}

%From this perspective, the language used is
%incidental (``mere choice of notation''), which helps explain
%the proliferation of languages --- any language that can call BLAS
%\cite{blas} will do!

%What do we mean by a \emph{language} for scientific computing?
%The prevailing
%answer is that it is about numeric arrays, perhaps matrices specifically,
%and numerical libraries that use them.
%Instead we intend to argue that, at a
%deeper level, it is about certain kinds of flexibility, particularly
%flexibility in the behavior of key operators and functions. This flexibility
%is needed to maximize composability and reusability in scientific code.
%Without it, certain programs may be easy to write today, but changing
%functional and performance requirements can become difficult to meet.

\section{The technical computing problem}

\begin{singlespace}
\begin{table}
  \begin{center}
    \def\arraystretch{1.25}
    \begin{tabular}{|l|l|}\hline
      \textbf{Mainstream PL} & \textbf{Technical computing} \\
      \hline \hline
      classes, single dispatch             &  complex operators \\
      \hline
      separate compilation                 &  performance, inlining \\
      \hline
      parametric polymorphism              &  ad-hoc polymorphism, extensibility \\
      % - concrete reuse of generated code
      \hline
      static checking                      &  experimental computing \\
      \hline
      modularity, encapsulation            &  large vocabularies \\
      \hline
      eliminating tags                     &  self-describing data, acceptance of tags \\
      \hline
      data hiding                          &  explicit memory layout \\
      \hline
    \end{tabular}
  \end{center}
  \caption[Programming language priorities]{
\small{
Priorities of mainstream object-oriented and functional programming language research and
implementation compared to those of the technical computing domain.
}
  }
  \label{PLpriorities}
\end{table}
\end{singlespace}

Table~\ref{PLpriorities} compares the general design priorities of mainstream
programming languages to those of technical computing languages.
The priorities in each row are not necessarily opposites or even mutually exclusive,
but rather are a matter of emphasis.
It is certainly possible to have both parametric and ad-hoc polymorphism within
the same language, but syntax, recommended idioms, and the design of the standard
library will tend to emphasize one or the other.
Of course, the features on the left side can also be useful for technical computing;
we exaggerate to help make the point.

It is striking how different these priorities are.
We believe these technical factors have contributed significantly to the persistence
of specialized environments in this area.
Imagine you want just the features on the left.
Then there are many good languages available: Haskell, ML, Java, C\#, perhaps C++.
Getting everything on the right, by comparison, seems to be awkward.
The most popular approach is to use multiple languages, as in e.g. NumPy~\cite{numpy},
with a high-level productivity language layered on top of libraries
written in lower-level languages.
Seeking a similar tradeoff, others have gone as far as writing a C++
interpreter \cite{vasilev2012cling}.

%technical computing systems have an unusually large amount of
%``failure of abstraction'' --- manual duplication of facts
%all over the system. imagine changing from column-major to
%row-major.

%problem of finding connections between array programming and OOP.
%array programming is a powerful paradigm for describing computational
%kernels operating over potentially large amounts of data.

%an ``object system'' in this context is often considered a separate
%part of the language, to be used only when arrays no longer
%suffice.

\section{Solution space}

It is clear that any future scientific computing language will need to be able to
match the performance of C, C++, and Fortran.
To do that, it is almost certain that speculative optimizations such as tracing
\cite{tracingjit} will not be sufficient ---
the language will need to be able to \emph{prove} facts about types, or at least
let the user request specific types.
It is also clear that such a language must strive for maximum convenience, or else
the split between performance languages and productivity languages will persist.

It is fair to say that two approaches to this problem are being tried: one is
to design better statically-typed languages, and the other is to apply
program analysis techniques to dynamically typed languages.
Static type systems are getting close to achieving the desired level
of flexibility (as in Fortress \cite{fortresspec} or Polyduce \cite{polyduce1},
for instance), but it is still too early to call a winner between these two
approaches (if, indeed, there even needs to be a winner).

Here we will use a dynamically typed approach, with approximate type inference
% TODO make a bigger deal out of Kell
(for an excellent discussion of the many meanings of the word ``type''
see \cite{Kell2014}).
There are several reasons for this.
First, insufficient static checking is most likely not the current limiting
factor in scientific computing productivity.
Second, some idioms in this domain appear to be inherently, or at least naturally,
dynamically typed (as we will explore in chapter~\ref{chap:2}).
Third, there has been a sense that ``dynamic'' or ``scripting'' language users do
not want to hear about types: they are thought of as a distraction and
associated with verbosity and nuisance compiler errors.
We hope to contribute an example of a language where types are useful and not
burdensome.

% making whatever high-level language faster leaves two things out:
% - providing more control for experts to request performance, or get
%   super high performance
% - making the language more flexible and powerful

% also for linking purposes the compiler will want to leave some
% record of what it has done, i.e. what assumptions it has made to
% generate the code it left behind.
% here you will want a type system.

%Finally, it is often the case that a seemingly-minor
%modification to a type system makes type checking undecidable. Rejecting
%such a system is totally reasonable. However if we are willing to
%accept undecidable checking from the beginning, opportunities arise to
%simplify and increase the power of the language.
%Occasionally one hears that ``dynamic type'' is an oxymoron, or that
%these types are not ``real types''

% mention value of having types that are agreed-on and work well within
% the language. enables so many optimizations.

%Efforts to analyze and optimize dynamically typed programs generally make two
%assumptions: (1) we should work on analyzing \emph{existing} popular
%languages, and (2) users of these languages don't want to use types.
%The first assumption makes practical sense.
%Convenience is hard to quantify, so using existing languages that have already
%been deemed convenient by popular opinion puts us on solid footing.
%This thesis describes a new language, so we simply take the negation of the
%first assumption as a premise.
%Our work addresses the second assumption more directly.
%We point out that types
%\footnote{}
%do not have to be used for static checking, and that using them for
%\emph{code selection} and \emph{code specialization} is particularly useful in
%technical computing.
%This perspective has not been explored thoroughly in the past.

% there is a kind of arms race between those who want more expressiveness
% and those trying to analyze and optimize programs. we propose a truce.

% one of our main innovations is a way to deftly navigate between two
% problems

\subsection{The excess power problem}

We claim that the amount of information that can be statically inferred
exceeds most dynamic languages' capacity to exploit it.
Much work has been done on program analysis and optimization techniques
for dynamically typed languages.
When static analyses (often incorporating run time information) are applied
to dynamically typed programs, it is typically possible to recover a
significant amount of type information (e.g. \cite{druby}).
What, then, can one do with this information?
If the goal is performance, various partial evaluations can be done:
generating code without type checks, removing branches, type-specializing
the storage of variables, and compile time method lookup are all valuable
and yield large real-world gains.

Method-at-a-time JIT compilers (e.g. \cite{grcevski2004java}
\cite{Suganuma:2005:DED:1075382.1075386})
can specialize method bodies on all arguments.
%and might use multiple dispatch \emph{internally} to select implementations
%at run time.
For example an optimizing compiler for Self~\cite{Dean:1994:TBI:182590.182489}
tracked all argument types per call site to help guide inlining.
Arguably, if method calls are dispatched on the first argument, but the types of all
arguments can be inferred, some power has been ``left on the table'' ---
we could have had multi-methods for little extra cost.
This argument does not apply equally to statically-typed languages, since they
cannot simply ``switch'' their functions to generic functions without significant
consequences for type checking.

This ``excess power'' problem applies to data structures as well.
For example, static or run time analysis might reveal that a certain array
can be represented as a native \texttt{Int32} array \cite{Bolz2013}.
If this information is not reflected in the source language, then
certain uses like passing data to native code become unnecessarily more
complicated.
And if one is going to implement homogeneous arrays anyway, why not
let programmers request them?
%Of course, a library like NumPy \emph{does} allow this, but with a
%mechanism

Some levels of performance are difficult to reach with implicitly
specialized code and data.
Given the knowledge that an array contains only \texttt{Int32} data,
we might want to go beyond essential optimizations like storing intermediate
values in registers, and actually use different algorithms.
For example, in Miller-Rabin primality testing~\cite{Rabin1980128}, checking
three ``witness'' values suffices for all 32-bit arguments, but up to seven
values might be needed for 64-bit arguments.
In cryptographic applications, exploiting this difference in an inner loop
could bring significant benefits.


\subsection{The divergence problem}

Another problem that occurs when analyzing programs with complex
type behavior is divergence: the analysis is likely to infer an
overly-large result from failing to eliminate enough possible
behaviors.
Narrowing the inferred types requires some extra source of type
information.
Multi-method signatures work well for this purpose.

past approaches include: optional user annotations, or a built-in
database of function type behavior, determined empirically or
analytically.
to solve the problem in general, library behavior should not be
built-in to the compiler.
however, when functions that are both highly complex and performance
critical are implemented at the library level it can be difficult
to ensure they are sufficiently optimized.
% specializer must decide when to widen, when to stop going around
% loops, etc.
% this is an opaque process.
% there is an essential divide between what it will statically
% evaluated, and what it won't.

However the divergence problem also places a limit on what
kinds of dispatch specifications are useful to program analyses:
some sets of values tend to be more robust under execution
than others.

\todo[inline]{Would love an example if possible}


Why do we prefer method signatures to imperative logic?

- it's extensible
- its declarative style is more compact
- allows writing functions in small pieces
  the more descriptive the types, the smaller the pieces
  might make compiler more efficient; removes large bits of code from
  consideration just from looking at types. don't need to analyze a
  large function just to discover that only a small part of it
  remains.
- clearer performance model. it's hard for programmer to guess
  when the compiler will statically evaluate an arbitrary expression
- allows defining semantics about things that are optimized, but
  *not* compile-time constants.
  method table can be (ab)used as a mutable lookup table structure.

% TODO:
% claim dispatch works better on approximate type info than branches
% when you have unionall types.
% dispatch is also extensible and declarative.

% doing analysis only, you can use arbitrarily complex types, but
% you are then limited by what the language can ``consume'', or
% by the sophistication of the t-functions you write.
% adding dispatch, the types become an abstraction under user
% control, so need to be meaningful.

%% what to dispatch on? dispatch power has been extended in many ways, but
%% there is no real limit to what somebody might want to dispatch on.
%% so what to do?
%% some sets of values are more robust under computation than others
%% (closure properties).
%% identify those sets using dataflow concerns.

%% say we have a method defined for integers, and also for the special cases
%% ``2'' and ``odd integers''. a realistic implementation
%% will group all of these under ``integer'', and ideally generate a couple
%% branches to handle the other cases. we argue the concept of ``integer''
%% here is a more robust set, and so a more fundamental language concept.

%- vs predicate dispatch: we extend dispatch power in a different direction,
%guided by semantic subtyping. goal is maximum power that still yields high
%likelihood of resolving many calls to a single implementation.


% on the ``sufficiently smart compiler''
% optimizing code is uncomputable in general, and in some sense
% AI-complete. as soon as a person names any *specific* optimization
% they want, it immediately becomes possible to imagine a compiler
% doing it.
% discussion on when dynamic typing is ``needed'' is related!
% again, isolated examples often unconvincing.
% same thing yet again tends to happen with multiple dispatch:
% it feels very useful overall, but many particular cases can be
% explained away as something single dispatch could handle.


\subsection{Staged programming}

In demanding applications, selecting the right algorithm might not
be enough, and we might need to automatically \emph{generate} code
to handle different situations.
While these cases are relatively rare in most kinds of programming,
they are remarkably common in technical computing.
Code generation, also known as \emph{staged programming}, raises
several complexities:

\vspace{-3ex}
\begin{singlespace}
\begin{itemize}
\item What is the input to the code generator?
\item When and how is code generation invoked?
\item How is the generated code incorporated into an application?
\end{itemize}
\end{singlespace}

\noindent
These questions lead to practical problems like ad-hoc custom syntax,
extra build steps, and excessive development effort (writing parsers and
code generators might take more work than the core problem).
If code needs to be generated based on run time information, then a
% TODO cite julia GPU paper that did this
staged library might need to implement its own dispatch mechanism
for invoking its generated code.

We find that our particular style of dynamic type-based dispatch provides
an unusually good substrate for staged programming.
The reason is that the language, though semantically dynamically typed,
has a canonical (and not optional!) type analysis stage.
This is a natural place to hook in code generators: semantic (not just
syntactic) information is available, but machine code has not been
generated yet.
In general, this semantic information can also include run time (and
always, at least, approximate run time) information.
Since the types used by dispatch have rich, nested structure, they
convey enough information to drive code generation for a significant
class of problems.
In this framework code generators are invoked automatically by
the compiler, and the results can be called directly or even inlined
into user code.
%A library can switched to a staged approach without affecting
%calling code.
Our approach is closely related to exotypes
\cite{DeVito:2014:FRG:2594291.2594307}.



%past (static) type systems for dispatch were designed to ensure the absence of
%no-method errors and ambiguities (completeness and uniqueness). our goal
%is instead to statically resolve methods. this is inherently heuristic and
%best-effort. since static types shouldn't affect program behavior, we
%conclude that the dispatch must be dynamic, which is happily the same
%conclusion you would reach if you simply wanted dynamic typing.

%it is quite possible that some static type system will work well for this
%however we defer this question.
%interesting variants:
%- require static single method matches
%- reject programs with no-method errors
%- reject programs that yield Unions

%%%%%%


%% The key ingredients:

%% \begin{singlespace}
%% \begin{enumerate}
%% \item Self-describing data model aware of memory layout
%% \item Type tags with nested structure
%% \item A fully-connected type tree
%% \item Dynamic multiple dispatch over all types % including parameters and varargs
%% \item Dataflow type inference
%% \item Automatic code specialization
%% \end{enumerate}
%% \end{singlespace}

%% This list of features may appear somewhat ad-hoc. However, they turn out to be
%% remarkably strongly coupled, and deeply constrained by our ultimate goal.
%% Each of these features has appeared in some form before, but never in a way
%% that fully solves the problems described here.

%% Challenges of this approach (why has this not been done before?)


\section{Contributions of this thesis}

% thesis stmt: integrating code selection and specialization with
% type-based dynamic dispatch captures both the performance and
% productivity requirements of technical computing.

% NUMBER ZERO

Our first contribution is a discussion of the nature of technical computing
that suggests which language-level abstractions might best support real use
cases.
Simply put technical computing has by and large happened
without sufficient probing and analysis as to what it is.
Our looking at the space in this manner has lead to a novel design.
The success of this design is resonating with users worldwide.

Based on the theory that technical computing is characterized by complex
operators and particular combinations of binding time behavior.

% NUMBER ONE

To our knowledge, the first technical computing system written in a single language.
Past systems have applied specialization to a set of user-level routines.
We show that it can be used for an entire system.

%sort of language would form a good base for it. it should emphasize
%complex operators and code generation/specialization.
% this includes our real-world experience with building a new
% technical computing environment from the ground up with a large open
% source community.

% NUMBER TWO

Second, we introduce the idea of integrated code selection and specialization.
%This leads to easy-to-use polymorphism, prioritizing
%flexibility and performance over nearly all else.

Using data flow types to drive the design leads to novel dispatch features:
- diagonal dispatch
- unifying variadic methods with regular types
- predictable performance without giving up flexibility

% Something about dispatch for performance. Given a performance kernel, it takes
% a lot of info to describe when it is applicable.

% dispatch on everything, not just classes

% more types lets you break up functions into more methods

% Instead of ``running dynamic code faster'', it's ``find the static part automatically''

% NUMBER THREE

Easy polymorphism

- library writers can decide what to put at the type level without affecting users
- a single mechanism to cover overloading, dispatch, and specialization
%- using types, but mostly for dispatch, where they can be used gradually
- specialize by default (makes static polymorphism less of a black art)


% NUMBER FOUR

Subtyping algorithm

% we design a type system for this. and introduce the ``evaluate softly
% and carry a big subtype relation'' school of thought.

% this subtype system is adoptable by dynamic languages that don't want to
% add checking, but might benefit from a richer abstract value domain in
% other ways.

% we provide a straightforward implementation that should be easy to
% translate to other languages.

% at this point, certain language implementation techniques have been fairly widely
% disseminated: how to write a parser, how to write a bytecode VM, how to
% design an object system, etc.
% subtyping systems are still relatively esoteric, often not even mentioned
% in introductory texts.

% NUMBER FIVE

Library design issues

James Morris eloquently observed that
``One of the most fruitful techniques of language analysis is explication through
elimination. The basic idea is that one explains a linguistic feature by showing
how one could do without it.'' \cite{morris}
An additional contribution of this thesis is the application of this approach to features
of technical computing environments that have not been subject to such analysis
before.
% showing our language is able to ``explain'' ...

% hopefully will provide useful examples for evaluating future dispatch
% systems, object systems, and type systems
